---
title: "Anaemia in pregnancy in Ghana"
subtitle: "Various analysis options"
author: "Shih-Ting Tseng and Proochista Ariana"
date: "`r format(Sys.time(), format = '%d %B %Y %H:%M')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  echo = FALSE
)

suppressPackageStartupMessages(source(here::here("packages.R")))
for (f in list.files(here::here("R"), full.names = TRUE)) source (f)

targets::tar_load(anc_data_recode)
```

In this document, we demonstrate steps taken in performing bivariate analysis of outcome variable (anaemia status) with available and potential exposure variables (age, profession, education level, marital status, location).

## Recode exposure variables to two category classifications

Currently, only the marital status variable uses a two category classification (married, single). We can simplify the other exposure variable such that they have a two category classification as well.

### Age

Bearing a child at an age less than 20 years old is considered high risk[^1],[^2] and are associated with various maternal morbidities (including anaemia) and maternal mortality. Given this, we can re-classify age into a two categories of those **20 years and older** and those **less than 20 years old**.

```{r}
## Processing code for re-classification of age ----
df <- anc_data_recode |>
  dplyr::mutate(
    early_childbearing = ifelse(age < 20, "Yes", "No") |>
      factor(levels = c("Yes", "No"))
  )

## Check re-classification ----
table(df$early_childbearing)
```

### Profession or employment or means of earning a living

Currently, the dataset contains 12 distinct professions or employment or means of earning a living. From the point of view of characterising those who are likely to suffer from anaemia during pregnancy, we can re-classify this to those who have a profession or employment or a means of livelihood and to those who do not.

```{r}
## Processing code for re-classification of profession ----
df <- df |>
  dplyr::mutate(
    livelihoods = ifelse(profession %in% c("None", "Student"), "No", "Yes") |>
      factor(levels = c("Yes", "No"))
  )

## Check re-classification ----
table(df$livelihoods)
```

### Education level

Currently, there are 4 educational level categories in the dataset based on the Ghanian education system. From the point of view of characterising those who are likely to suffer from anaemia during pregnancy, we can re-classify this to those who have had any formal education and to those who have not. This re-classification assumes that even primary education is a distinguishing feature that can differentiate women who suffer from anaemia during pregnancy. However, literature suggests that it might be about secondary education and higher that is more important[^3].

```{r}
## Processing code for re-classification of education level ----
df <- df |>
  dplyr::mutate(
    higher_education = ifelse(education_level %in% c("None", "Primary"), "No", "Yes") |>
      factor(levels = c("Yes", "No"))
  )

## Check re-classification ----
table(df$higher_education)
```

### Location

Currently, we have information on the town/general area where the pregnant women are from. This exposure variable proxies distance from the clinic. In similar studies done in the context of Ghana[^4], distance to services was classified as being services within community or outside community. This is a possible classification that can be used with our dataset. The main thing to determine is a criteria for classifying within community services. This can be a distance criteria using the health clinic catchment area concept. Generally, clinic catchment area is defined as the area up to 10 kms radius from the health clinic.

Another way of approaching this is using the dataset used in the paper on spatial inequalities in skilled birth attendance in Ghana[^5] and then plotting the geolocations of the towns and of the health clinic and then calculating a rough distance between these towns and the health clinic. Anything that is more than 10 kms in distance will be considered outside community and anything 10 kms or less will be considered within community


[^1]: https://data.unicef.org/topic/child-health/adolescent-health/#:~:text=Early%20pregnancy%20and%20childbearing%20can,and%20early%20and%20forced%20marriage.

[^2]: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8173845/#:~:text=Girls%20who%20get%20married%20and,been%20quantified%20in%20previous%20studies.

[^3]: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5954959/#:~:text=3.3.&text=Women%20with%20secondary%20education%20or,associated%20with%20anaemia%20during%20pregnancy.

[^4]: https://link.springer.com/article/10.1007/s43999-024-00042-0

[^5]: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7613541/